---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r warning=FALSE}
# Load the necessary libraries
library(ggplot2)
suppressMessages(suppressWarnings(library(gridExtra)))
suppressMessages(suppressWarnings(library(dplyr)))
# Read the data 
activity <- read.table("activity.csv",
                       header=TRUE,quote="\"", sep=",",
                       colClasses=c("integer","Date","integer"))

```
Summarize the data, to assess whether there are missing values, range and number of values.
```{r}
str(activity)
summary(activity$steps)
```
Ploting a histogram of the number of steps per day. The aggregate function does not return any day which has all NA values.

```{r warning=FALSE}
day_total <- aggregate(steps ~  date, data=activity, sum)
ggplot(day_total,aes(steps)) + 
      geom_histogram(binwidth=500,fill="blue",colour="black") + 
      xlab("Number of Steps in a day") + 
      ylab("Number of days") + 
      ylim(0,8) +
      ggtitle("Histogram of number of steps per day")
```

## What is mean total number of steps taken per day?
The mean total number of steps taken per day is `r format(mean(day_total$steps),digits=5)` steps. The median total number of steps taken per day is `r format(median(day_total$steps),digits=5)` steps


## What is the average daily activity pattern?
Use the aggregate function to find the mean for each interval (0000 to 2355).
```{r}
interval_mean <- aggregate(steps ~ interval, data=activity, mean)
ggplot(interval_mean,aes(interval,steps)) + 
      geom_line(color="blue") + 
      xlab("Interval, hhmm") +
      ylab("Average number of steps") +
      ylim(0,250) +
      ggtitle("Average number of steps per interval")
```
The maximum number of steps occurs in the interval `r interval_mean[which.max(interval_mean$steps),1] ` 

## Imputing missing values

The number of rows with NA values is `r sum(is.na(activity$steps))`. A simple approach for imputing missing values is to replace any interval NA values with the average value for that interval across all the measurements. 
``` {r}
activity_no_NA <- merge(x=activity, y=interval_mean, by="interval")
activity_no_NA <- mutate(activity_no_NA, steps = ifelse(is.na(steps.x)==TRUE,steps.y,steps.x))
# Drop the steps.x and steps.y columns
activity_no_NA <- subset(activity_no_NA,select = -c(steps.x,steps.y))
activity_no_NA <- activity_no_NA[order(activity_no_NA$date,activity_no_NA$interval),]

day_total_no_NA <- aggregate(steps ~  date, data=activity_no_NA, sum)
plot1<-ggplot() + 
      geom_histogram(data=day_total,aes(steps),binwidth=500,fill="blue",colour="black") +
      xlab("Number of Steps in a day") + 
      ylab("Number of days") +ylim(0,15) +ggtitle("Comparision of with NA and with imputing of NA values")

plot2<- ggplot() + 
      geom_histogram(data=day_total_no_NA,aes(steps),binwidth=500,fill="red",colour="black") +
      xlab("Number of Steps in a day") + 
      ylab("Number of days") +ylim(0,15)

 grid.arrange( plot1, plot2, nrow=2)

```
By observation, the bin centred on 11000 has increased its count by 8 days, representing the number of days that were filled with NA values.
For the data set with the NA values replaced with the interval's mean value, the mean =  `r format(mean(day_total_no_NA$steps),digits=5)` (was `r format(median(day_total_no_NA$steps),digits=5)` ) and the median value is `r format(mean(day_total$steps),digits=5)` (was `r format(median(day_total$steps),digits=5)` ).
There is little change to these statistics through the replacement of NA with the mean value for each interval.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Assign the day of week based on the the measurement was taken
activity_no_NA$day_of_week <- weekdays(activity_no_NA$date)
# assign the grouping,day_type for 'weekend' if the day of week is Saturday or Sunday; everything else is a week day
 activity_no_NA$day_type <- ifelse(activity_no_NA$day_of_week=="Sunday"|activity_no_NA$day_of_week=="Saturday","weekend","weekday")
 # Add the day_type as a factor to the data.frame
 activity_no_NA$day_type <- as.factor(activity_no_NA$day_type)
 # Aggregate based on interval, mean of steps and  day_type as a factor.
 interval_mean_day_type <- aggregate(cbind(steps) ~ interval + day_type, data=activity_no_NA, mean)
 ggplot(interval_mean_day_type,aes(x=interval,y=steps)) +
       geom_line(color="blue") + 
       facet_grid(day_type~.) +
       ggtitle("Comparision of Weekend and Weekday steps per interval") +
   xlab("Interval, hhmm") + 
      ylab("Average number of Steps")
 
 ```
The two line plots show that the subject has a higher, more compressed steps per interval early in the day for the weekdays. The number of steps per interval for the weekends has a larger spread compared to the weekdays graph.
 